---
title: "ASCI 944 / STAT 844 Quantitative Methods for Genomics of Complex Traits"
subtitle: "Homework assignment 6"
author: "Gota Morota"
date: "April 12, 2018"
output:
  rmdformats::readthedown:
    highlight: kate
    self_contained: false
    code_folding: hide
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

### Due date
Thursday, April 19, 5pm

## Data 
This homework assignment involves data analysis of 280 winter wheat accessions genotyped with 1083 Diversity Array Technology (DArT) markers at 9 locations. Phenotypes available are grain yield, grain volume weight, plant height, and flowering date. Both phenotypic and genotypic data are downloadable from [DRYAD](http://datadryad.org/resource/doi:10.5061/dryad.2sk59). 
```{r, eval=FALSE}
# phenotypes 
pheno <- read.csv("http://datadryad.org/bitstream/handle/10255/dryad.40880/phenotype.csv?sequence=1", header = TRUE, stringsAsFactors = FALSE)
pheno1 <- subset(pheno, Location ==1) # use location 1

# DArt markers 
library(readxl)
geno1 <- read_excel("DartGenot.xls", col_names=TRUE, sheet=2, skip=3, na="-")
geno2 <- read_excel("DartGenot.xls", col_names=TRUE, sheet=3, skip=3, na="-")
geno3 <- read_excel("DartGenot.xls", col_names=TRUE, sheet=4, skip=3, na="-")
table(geno1[,1:7] == geno2[,1:7]) 
table(geno1[,1:7] == geno3[,1:7])
table(geno2[,1:7] == geno3[,1:7])
geno4 <- cbind(geno1, geno2[,-c(1:7)], geno3[, -c(1:7)])
```

## Quality control
```{r, eval=FALSE}
# quality control
qc.callrate <- which(geno4$CallRate < 95)
qc.p <- which(geno4$P < 80)
qc.index <- unique(c(qc.callrate, qc.p))
length(qc.index) # 1234
geno5 <- geno4[-qc.index, ]
geno6 <- t(geno5[,-c(1:7)])
```

## Data cleaning 
Check if the phenotype and genotype files have the same accessions. 
```{r, eval=FALSE}
# phenotype -> genotype
table(pheno1[,2] %in% rownames(geno6))
na1a.index <- which(!pheno1[,2] %in% rownames(geno6))
pheno1[c(na1a.index),2] # "NE10522" "NW10568" "NE10570" "NE10583" -> found in the phenotype file but in the genotype file 

# genotype -> phenotype 
table(rownames(geno6) %in% pheno1[,2])
na1b.index <- which(!rownames(geno6) %in% pheno1[,2])
rownames(geno6)[c(na1b.index)] # "Goodstreak" "Camelot" -> found in the genotype file but in the phenotype file 

pheno1a <- pheno1[match(rownames(geno6), pheno1[,2]), ]
pheno1b <- pheno1a[-c(277:278), ]
geno7 <- geno6[-c(277:278), ]
table(pheno1b[,2] == rownames(geno7), useNA = "always")
table(pheno1b[,2] %in% rownames(geno7))
table(rownames(geno7) %in% pheno1b[,2])

# final phenotype object 
y <- pheno1b$yield # use grain yield 

# final genotype object 
geno <- geno7 # 276 x 747
```

## Question 1
Replace missing marker genotypes with mean values. Then store the marker genotypes in a matrix object `X`.
```{r, eval=FALSE, echo=FALSE}
X <- matrix(0, ncol=ncol(geno), nrow=nrow(geno))
for (j in 1:ncol(geno)){
  #cat("j = ", j, '\n')
  X[,j] <- ifelse(is.na(geno[,j]), mean(geno[,j], na.rm=TRUE), geno[,j])
}
```

## Question 2
Perform a quality control by removing markers with MAF < 0.05. How many markers are removed? Save the filtered genotype matrix in `X2`.
```{r, eval=FALSE, echo=FALSE}
freq <- colMeans(X)
maf <- ifelse(freq > 0.5, 1-freq, freq)
maf.index <- which(maf < 0.05)
length(maf.index)
X2 <- X[, -maf.index]
dim(X2)
```

## Question 3
Standardize the genotype matrix from Question 2 to have a mean of zero and variance of one. Save this matrix as `Xs`.
```{r, eval=FALSE, echo=FALSE}
Xs <- scale(X2, center = TRUE, scale = TRUE)
```

## Question 4
We will determine a prior for marker effects in Bayesian ridge regression. Read Perez and de los Campos (2014) ([10.1534/genetics.114.164442](www.dx.doi.org/10.1534/genetics.114.164442)) to learn more about the BGLR R package. 
Recall that genetic variance is given by $Var(g_i) = \sigma^2_b \sum^m_{j=1} Var(x_{ij})$. We can equate this $\sigma^2_b \times \sum^m_{j=1} Var(x_{ij})$ to the product of our prior expectation about the expected proportion of variance that is explained by the regression times an estimate of the phenotypic variance  $\sigma^2_b \times \sum^m_{j=1} Var(x_{ij}) = R^2 V_y$. This results in $\sigma^2_b  = \frac{R^2 V_y}{\sum^m_{j=1} Var(x_{ij})}$. 
Then equating this to the prior mode of the variance parameter of scaled-inverse chi square density gives $\frac{R^2 V_y}{\sum^m_{j=1} Var(x_{ij})} =  \frac{S_b}{df_b + 2}$, where $S_b$ and $df_b$ are the prior scale and degrees of freedom of marker effects, respectively. Then solving for the prior scale parameter $S_b$ yields $S_b =  \frac{R^2 V_y}{\sum^m_{j=1} Var(x_{ij})} (df_b + 2)$. Compute the prior scale $S_b$ for the wheat data set according to the above derivation. Use $df_b = 5$ and $R^2 = 0.5$.
```{r, eval=FALSE, echo=FALSE}
# prior for Baysian ridge regression - \sum (Var(\sum^m_{j=1}x_{ij} ) )
x2 <- sum(apply(X=Xs, MARGIN = 2, FUN = var))
MSx <- x2
R2 <- 0.5
(R2 * var(y) * (5+2)) / MSx # 0.3253088
```

## Question 5
Alternatively, we can use $Var(g_i) = \sigma^2_b \times n^{-1} \sum^n_{i=1} \sum^m_{j=1}x^2_{ij}$, where $n^{-1} \sum^n_{i=1} \sum^m_{j=1}x_{ij}^2$ is the average sum of squares of the genotypes. Recompute the prior scale $S_b$ based on this parameterization. Again, use $df_b = 5$ and $R^2 = 0.5$.
```{r, eval=FALSE, echo=FALSE}
# prior for Bayesian ridge regression - \sum^n_{i=1}\sum^m_{j=1}x^2_{ij} / n
x2 <- sum(apply(X = Xs, MARGIN = 2, function(x) sum(x^2))) # sum(Xs*Xs)
x2mean = sum((apply(X = Xs, MARGIN = 2, mean))^2)
MSx <- x2/nrow(Xs) - x2mean
R2 <- 0.5
(R2 * var(y) * (5+2)) / MSx # 0.326491
```

## Question 6
Report a prior scale $S_b$ used for Bayesian ridge regression with default setting in the BGLR function. Which of the above rule-based priors is used in the BGLR? Hint: set `nIter` = 1, `burnIn` = 1 so that you can see the output. 
```{r, eval=FALSE, echo=FALSE}
library(BGLR)
# rule-based prior 
ETA <- list(MRK=list(X = Xs, model="BRR"))
fit.BRR <- BGLR(y = y, ETA = ETA, nIter = 1, burnIn = 1)
```


## Question 7
Evaluate predictive performance of the Bayesian ridge regression model by repeating 3-fold cross-validation 5 times. Use `set.seed(0403)` at the begining of the code so that your analysis is reporducible. 
```{r, eval=FALSE, echo=FALSE}
library(BGLR)
set.seed(0403)
r2 <- array()
for (i in 1:5){
  folds <- 3
  partition <- sample(1:3, size=nrow(Xs), replace = TRUE)
  r1 <- array()
  for (j in 1:folds){
    y.na <- y
    tst <- which(partition == j)
    y.na[tst] <- NA
    ETA <- list(MRK=list(X = Xs, model="BRR"))
    fit.BRR <- BGLR(y = y.na, ETA = ETA, nIter = 5000, burnIn = 2000, verbose = FALSE)
    r1[j] <- cor(y[tst], fit.BRR$yHat[tst])
  }
  r2[i] <- mean(r1)
}
mean(r2) 
```


## Question 8
Evaluate predictive performance of the Bayesian LASSO regression model by repeating 3-fold cross-validation 5 times. Use `set.seed(0403)` at the begining of the code so that your analysis is reporducible. Compare predictive accuracies between the Bayesian ridge regression and Bayesian LASSO. 
```{r, eval=FALSE, echo=FALSE}
library(BGLR)
set.seed(0403)
r2 <- array()
for (i in 1:5){
  folds <- 3
  partition <- sample(1:3, size=nrow(Xs), replace = TRUE)
  r1 <- array()
  for (j in 1:folds){
    y.na <- y
    tst <- which(partition == j)
    y.na[tst] <- NA
    ETA <- list(MRK=list(X = Xs, model="BL"))
    fit.BRR <- BGLR(y = y.na, ETA = ETA, nIter = 5000, burnIn = 2000, verbose = FALSE)
    r1[j] <- cor(y[tst], fit.BRR$yHat[tst])
  }
  r2[i] <- mean(r1)
}
mean(r2) 
```

