---
title: "ASCI 944 / STAT 844 Quantitative Methods for Genomics of Complex Traits"
output: 
  BiocStyle::html_document
---

# Gaussian kernel 
Ornella et al. (2014) ([doi:10.1038/hdy.2013.144](http://dx.doi.org/10.1038/hdy.2013.144)) analyzed 14 and 16 trait–environment combinations in maize and wheat, respectively. You can download the wheat stem rust—main season data set in `Wheat-Stem Rust.zip` from the [CIMMYT data repository](http://repository.cimmyt.org/xmlui/handle/10883/2976). 

## Data loading (X $\in (0, 1)$)
```{r}
load("StemRust/SR_F6PAV0N_Srm.RData") 
```

## Impute missing genotypes by mean values 
```{r}
X <- matrix(0, ncol=ncol(Markers), nrow=nrow(Markers))
for (j in 1:ncol(Markers)){
   X[,j] <- ifelse(is.na(Markers[,j]), mean(Markers[,j], na.rm=TRUE), Markers[,j])
}
```

## Remove markers with MAF < 0.01
```{r}
freq <- colMeans(X) # X \in (0,1)
maf <- ifelse(freq > 0.5, 1-freq, freq)
maf.index <- which(maf < 0.01)
length(maf.index)
X <- X[, -maf.index] #  176 x 931
```

## Define G matrix and Gaussian kernel functions
```{r}
# G matrix
computeG <- function(X){
  X <- scale(X, center = TRUE, scale = TRUE)
  p <- ncol(X)
  G <- tcrossprod(X) / p
  return(G)
}

# Gaussian kernel matrix
computeGK <- function(X, bw){
  X <- scale(X, center = TRUE, scale = TRUE)
  D<- as.matrix(dist(X, method="euclidean"))^2
  D <- D/mean(D) # or just use D
  GK <- exp(-bw * D) 
  return(GK)
}
```

## Comparison 1
```{r}
# G matrix
G <- computeG(X)
hist(G[upper.tri(G)], main="", xlab="G(i,i')")

# GK with theta = 1.0
GK1.0 <- computeGK(X, bw=1.0)
hist(GK1.0[upper.tri(GK1.0)], main="", xlab="K(i,i')")
title(substitute(theta == 1.0))

# GK with theta = 0.7
GK0.7 <- computeGK(X, bw=0.7)
hist(GK0.7[upper.tri(GK0.7)], main="", xlab="K(i,i')")
title(substitute(theta == 0.7))

# GK with theta = 0.1
GK0.1 <- computeGK(X, bw=0.1)
hist(GK0.1[upper.tri(GK0.1)], main="", xlab="K(i,i')")
title(substitute(theta == 0.1))
```


## Comparison 2 
```{r}
par(mfrow=c(2,2))
G <- computeG(X)
hist(G[upper.tri(G)], main="", xlab="G(i,i')")
GK1.0 <- computeGK(X, bw=1.0)
hist(GK1.0[upper.tri(GK1.0)], main="", xlab="K(i,i')")
title(substitute(theta == 1.0))
GK0.7 <- computeGK(X, bw=0.7)
hist(GK0.7[upper.tri(GK0.7)], main="", xlab="K(i,i')")
title(substitute(theta == 0.7))
GK0.1 <- computeGK(X, bw=0.1)
hist(GK0.1[upper.tri(GK0.1)], main="", xlab="K(i,i')")
title(substitute(theta == 0.1))
dev.off()
```

## Multi-kernel RKHS regression
```{r}
library(BGLR)
set.seed(0728)
tst <- sample(1:nrow(X), size=76)
y.na <- y
y.na[tst] <- NA

ETA <- list(K1=list(K = GK1.0, model="RKHS"), K2=list(K = GK0.1, model="RKHS"))
fit.RKHS <- BGLR(y = y.na, ETA = ETA, nIter = 5000, burnIn = 2000, verbose = FALSE)
cor(y[tst], fit.RKHS$yHat[tst])
```

